---
title: 'Data 608: Final Project'
author: "Christina Valore"
date: "12/15/2019"
output: 
  html_document:
    #df_print: paged
    toc: true # table of content true
    toc_collapsed : false
    toc_float: true
    code_folding : show
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    smooth_scroll: true
    theme: darkly  # many options for theme, this one is my favorite.
    highlight: textmate  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
library(dplyr);library(ggplot2)
library(ggthemes); library(e1071); library(caTools); library(MASS); library(funModeling); library (plyr)
```

# Overview

For this project, I investigated the on-time performance (OTP) of NYC subways as I am a commuter in NYC and I have always wondered what years/months the subways performed the worst. From experience, I would guess that during the winter/fall seasons, the subways tend to be less on-time.

For the analysis, MTA subway data was used from the years 2011-2017 during the weekdays. Our business question was: *What year was the MTA subway most likely to be late?* After acquiring and cleaning the data, I then built visualizations using ggplot and plotly. 

Our findings showed that 2017 was the worst performing year with months like December and November being the worst months to ride the subway. 

# About the Data

Data was obtained from the MTA website: http://web.mta.info/developers/performance.html

For the *MTA data*, it was first uploaded directly into GitHub in a .CSV and then into R. The data was then subsetted to pull out only the OTP (on time performance) subway data and stored in an R dataframe. I also changed the integer months to strings.

* About the MTA subway performance dataset:

  *  Subway wait assessment for all lines
  *  Customer injury rate
  *  Elevator availability
  *  Escalator availability
  *  Mean distance between failure
  *  On-time performance (OTP), total and for all lines
  *  Total ridership

```{r MTA}
# load MTA data 
data<- read.csv("https://raw.githubusercontent.com/ChristinaValore/stats-prob-606/master/Performance_NYCT.csv", header=TRUE, check.names = FALSE)

# subset to pull out ONLY the OTP (ON TIME PERFORMANCE)
sub <-data[ which(data$PARENT_SEQ=='391690'), ]

# view the subsetted data
head(sub,10)

# remove columns that are not needed for this analysis
sub<- sub[c(-1:-3, -5:-10)]

# change int months to string months
sub$PERIOD_MONTH<-with(sub, month.abb[PERIOD_MONTH])
```

# Analysis

I first checked for the completeness of the data as some years were missing OTP values and some subway lines only had collected data from the mid-year. We removed all incomplete data including any years after 2011 and incomplete OTP for the subway lines W and S Line 42 St.

After, a Boolean comparison was done on the monthly actual vs. monthly target OTP and a new column, ON_TIME, was created with {1,0} values to indicate if the subway was on-time for that month.

Finally, the percentage values for OTP was changed to decimals, divided by approximately 20 days as only weekdays are considered in the collected data and added into a new column called DAYS_ON_TIME. I also did a quick check again on the Na's and O values to be sure there was no missing data remaining (excluding the variable ON_TIME as that can only be {1,0}) 

The new cleaned set was then export to .CSV to use in the shiny app. I also created two ggplots below using a facet wrap. 

```{r subway cleanse}
# check for data completeness to see how many subways were analyzed
unique(sub$INDICATOR_NAME)

# 24 subways were analyzed over 12 months so each year should have 288 values
table(sub$PERIOD_YEAR) 

# See if OTP was measured for all subway lines
table(sub$INDICATOR_NAME)

# remove the years 2009 and 2010 and standardize the remainder of the years be removing incomplete subway lines: S line 42 & the W line
sub.updated<-sub %>% 
  filter(PERIOD_YEAR >= "2011", INDICATOR_NAME != "OTP (Terminal) - W Line", INDICATOR_NAME != "OTP (Terminal) - S Line 42 St.")

# check on the data once more
table(sub.updated$PERIOD_YEAR) 
table(sub.updated$PERIOD_MONTH) 

#change line names by removing OTP (Terminal)
sub.updated$INDICATOR_NAME<-revalue(sub.updated$INDICATOR_NAME, c("OTP (Terminal) - 1 Line" = "one" , "OTP (Terminal) - 2 Line"="two", "OTP (Terminal) - 3 Line"="three","OTP (Terminal) - 4 Line"="four","OTP (Terminal) - 5 Line"="five","OTP (Terminal) - 6 Line"="six","OTP (Terminal) - 7 Line"="seven","OTP (Terminal) - A Line"="A","OTP (Terminal) - B Line"="B","OTP (Terminal) - C Line"="C","OTP (Terminal) - E Line"="E","OTP (Terminal) - F Line"="F","OTP (Terminal) - D Line"="D","OTP (Terminal) - G Line"="G", "OTP (Terminal) - J Z Line" = "J Z","OTP (Terminal) - L Line"="L","OTP (Terminal) - M Line"="M","OTP (Terminal) - N Line"="N","OTP (Terminal) - Q Line"="Q","OTP (Terminal) - R Line"="R","OTP (Terminal) - S Fkln Line" = "S Fkln","OTP (Terminal) - S Line 42 St." = "S 42 St.","OTP (Terminal) - S Line Rock" = "S Rock", "OTP (Terminal) - W Line" = "W"))

```

```{r}
# change the structure of certain variables to factors
sub.updated$PERIOD_YEAR<-as.factor(sub.updated$PERIOD_YEAR)
sub.updated$PERIOD_MONTH<-as.factor(sub.updated$PERIOD_MONTH)

# compare the monthly actual OTP to the monthly target OTP and if the actual is >= then the value is true
sub.updated$ON_TIME <- as.numeric(sub.updated$MONTHLY_ACTUAL >= sub.updated$MONTHLY_TARGET)

# change percentages to decimals
sub.updated$MONTHLY_TARGET <-sub.updated$MONTHLY_TARGET/100
sub.updated$MONTHLY_ACTUAL <-sub.updated$MONTHLY_ACTUAL/100
sub.updated$YTD_TARGET <-sub.updated$YTD_TARGET/100
sub.updated$YTD_ACTUAL <-sub.updated$YTD_ACTUAL/100

# approximately 20 weekdays in a month as the OTP is only measured for weekdays
sub.updated$DAYS_ON_TIME<-sub.updated$MONTHLY_ACTUAL*20

#check for any missing values
df_status(sub.updated, print_results = TRUE)
```

```{r}
#export sub updated to .CSV to use the cleaned version in app 
write.csv(sub.updated,'C:\\users\\christinavalore\\Desktop\\MTA_CLEANED.csv', row.names = FALSE)
```

# Visualizations

Using ggplot2, we facet wrap the month vs. days on time with the year and see that after 2013, the days on time start to decrease, with the worst OTP in 2017. 

```{r}
#facet wrap by year
qplot(PERIOD_MONTH, DAYS_ON_TIME, data = sub.updated, facets = . ~ PERIOD_YEAR) + theme(axis.text.x = element_text(size = 5, angle = 90))
```

I then facet wrap the month vs days on tine with the subway line and see that the worst OTP comes from the 2 and 4/5 line. 

```{r}
#facet wrap by year
qplot(PERIOD_MONTH, DAYS_ON_TIME, data = sub.updated, facets = . ~ INDICATOR_NAME) + theme(axis.text.x = element_text(size = 3, angle = 90))
```

# Shiny App

Using the shiny app functionality, I then take the cleaned dataset and use dpylr to filter the date out by year and month to see how the subway lines perform. 

To see the subway OTP over the years in the month of November: https://cvalore.shinyapps.io/OTP_MONTH/

To see the subway OTP over the months for 2017: https://cvalore.shinyapps.io/OTP_YEAR/

I chose the year 2017 first as from the ggplot above it looked as if that was the poorest performing year. By examining the app, I could see December was often at the end of the graphs meaning it was the poorest scoring month.

So then I created an app that targeted just December over the years. 

# Conclusion

From my visualizations I would say that over the years the on-time performance for subways became poorer, specifically in the year 2017. As I imagined the winter months, like December and even at times November, showed poor on-time performance. 

I did a quick Google search to see if I could find any evidence to back my claims up as to say that 2017 was a low performing year and I found this article: https://NY.curbed.com/2017/6/14/15801694/mta-NYC-subway-delays-twitter titled: MTA WTF: A visual timeline of the subway’s epic 2017 meltdown.

Although the article is just stating how bad the MTA infrastructure and delays were in 2017, sighting only a handful examples, it is still interesting to see how this was something that was quite noticeable to NYC during this year. So noticeable that on Jun 29, 2017, the Governor declared a state of emergency for the NYC Subway!


